NGO Repression as a Predictor of Worsening Human Rights Abuses

### 2021-11-17 manuscript/manuscript.Rmd
Suparna edits

---
title: "NGO Repression as a Predictor of Worsening Human Rights Abuses"
short-title: "NGO Repression and Human Rights"
author:
- name: Suparna Chaudhry
  affiliation: Lewis & Clark College
  email: schaudhry@lclark.edu
  url: http://www.suparnachaudhry.com/
- name: Andrew Heiss
  affiliation: Georgia State University
  email: aheiss@gsu.edu
  url: https://www.andrewheiss.com/
date: "November 15, 2021"
published: "Conditionally accepted at the *Journal of Human Rights*"
code-repo: "Access the code and full analysis notebook at <https://stats.andrewheiss.com/canary-ngos/>"
abstract: >-
  An increasing number of countries have recently cracked down on non-governmental organizations (NGOs). Much of this crackdown is sanctioned by law and represents a bureaucratic form of repression that could indicate more severe human rights abuses in the future. This is especially the case for democracies, which unlike autocracies, may not aggressively attack civic space. We explore if crackdowns on NGOs predict broader human rights repression. Anti-NGO laws are among the most subtle means of repression and attract lesser domestic and international condemnation compared to the use of violence. Using original data on NGO repression, we test whether NGO crackdown is a predictor of political terror, and violations of physical integrity rights and civil liberties. We find that while formal de jure laws provide little information in predicting future repression, their patterns of implementation, or de facto civil society repression, predicts worsening respect for physical integrity rights and civil liberties.
# keywords: civil society, NGOs, repression, human rights, political terror, civil liberties
word-count: "8,007"
thanks: >-
  We would like to thank participants of the 2021 International Studies Association (ISA) convention, and the "Human Rights on the Edge: The Future of International Human Rights Law and Practice" workshop hosted by the Global Human Rights Hub at Arizona State University for their comments and feedback.
reference-section-title: References
bibliography: bibliography.bib
toc: false
title-page: true
endnotes: true
endfloat: false
mainfont: Cochineal
sansfont: Inter
author-note: >-
  Suparna Chaudhry, PhD, Assistant Professor of International Affairs, Department of International Affairs, Lewis \& Clark College; Andrew Heiss, PhD, Assistant Professor of Public Management, Andrew Young School of Policy Studies, Georgia State University. 
correspondence: >-
  Correspondence concerning this article should be addressed to Andrew Heiss, PhD, Department of Public Management and Policy, Andrew Young School of Policy Studies, Georgia State University, 14 Marietta Street NW, Suite 357, Atlanta, GA 30303. Contact: aheiss@gsu.edu
editor_options: 
  chunk_output_type: console
---

@@

Over the last few decades, governments across the globe have been increasingly cracking down on civil society organizations (CSOs), particularly non-governmental organizations (NGOs)—a phenomenon commonly known as closing or shrinking civic space [@CarothersBrechenmacher:2014; @DPR:2016; @Chaudhry:2016; @DupuyPrakash:2018]. More than 100 countries have proposed or enacted 244 different measures that have restricted, repressed, or shut down civil society since 2013 [@icnl:2021]. In Human Rights Watch's 2016 World Report, Executive Director Kenneth Roth argued that civil society was under more aggressive attack than at any time in recent memory, and over the past decade, the majority of the world has seen a substantial narrowing of allowable civic space. As seen in Figure \@ref(fig:civicus-map), by the end of 2020, only `r civicus$Open$prop_nice` of countries had open and unrestricted civic space [@civicusmonitor; @ChaudhryHeiss:2021].[^open]

@@

Much of this targeting of NGOs has been non-violent. Rather than publicly arrest or beat activists, states implement "administrative crackdown," or the passage of laws to create barriers to NGO advocacy, entry, and funding [@Chaudhry:2016]. The use of legal tools instead of violent repression means that NGOs have to adapt their practices to stay within the bounds of permitted activities and operations [@Heiss:2019; @Heiss:2017]. While this can lead to the depoliticization and avoidance of direct political activism by CSOs [@BloodgoodTremblay-Boire:2017], states cannot soften or neutralize all forms of political activity and advocacy in subtle, legal ways like they have with NGOs. Sooner or later, states may need to deal with threats to their stability in a more direct, violent, confrontational manner. Anti-NGO laws could be an important early warning sign—or a canary in the coal mine—of worsening human rights repression, signaling that a state will continue down the path of further repression. What does state targeting of NGOs tell us about the state's broader respect for civil society and the overall level of human rights repression? Can the crackdown on NGOs act as a predictor of subsequent increasing repression?

In this paper, we explore the power of anti-NGO repression in predicting political terror and underlying human rights abuses. We argue that information about NGO crackdown can be a valuable indicator of human rights trajectories. It can act as an early warning signal for worsening repression, suggesting which countries are testing the waters with comparatively "milder" methods of crackdown before enacting more violent forms of repression. To assess if attacks on NGOs and activists can provide us useful information about future repression, we use an original dataset on state crackdown on NGOs using legal repression across all countries from 1990–2013. We differentiate between two forms of NGO repression: official de jure anti-NGO restrictions that are codified in law, and their de facto implementation. Using multilevel Bayesian modeling, we find that formal de jure laws potentially provide a weak signal of future repression. However, we find that accounting for their implementation does boost our models' predictive power. Increasing de facto civil society repression, on the other hand, predicts both a moderate increase in the probability of moderate political terror and a deterioration in the overall respect for human rights in a typical country.

Our paper makes an important contribution in predicting the onset of repressive state policies. While structural factors such as levels of democracy, economic development, electoral competitiveness can identify countries that are likely to have more or less respect for civic freedoms, these factors are also slow to change over time. As such, they may not be particularly suitable for forecasting short-term trajectories, especially for moderately repressive countries. On the other hand, state repression of NGOs is a more dynamic process that may provide insight into changing levels of human rights standards and repression in subsequent years. While recent research has made important contributions to studying the extent and effects of state crackdown on NGOs, much of this focuses on de jure restrictions and its impact on NGOs and donors [@christensen2013; @DPR:2016; @DupuyPrakash:2018; @BromleySchoferLonghofer:2020; @BakkeMitchellSmidt:2020; @SmidtPereraMitchell:2021]. By looking at laws *and* their implementation, as well as the extra-legal treatment of NGOs, we are able to assess the long-term effects of closing civic space on broader patterns of repression.

@@

Below, we first look at existing literature on repression and political violence, outlining state motivations to repress NGOs. We then elaborate on why attacking NGOs may be an important predictor of subsequent repression. Using both original data as well as data from several existing datasets, we build an ensemble of predictive models and explore (1) the marginal effects of civil society repression on different forms of broader human rights repression, and (2) the out-of-sample predictive performance of each model. We then illustrate one of the mechanisms of our findings—the disconnect between de jure and de facto anti-NGO repression—with a brief case study of Egypt's treatment of civil society since 2002. We conclude by discussing the implications of our findings on autocratic survival, and how states may manage NGOs as a strategy to remain in power.

@@

Repression is the use of coercive action against an individual or organization, within the territorial jurisdiction of the state, for the purpose of imposing a cost on the target as well as deterring specific activities perceived to be challenging to government personnel, practices, or institutions [@Goldstein:1978, p. xxvii; @Davenport:2007, 487; and @ConradRitter:2019, 15]. Early literature on repression focused on the role of structural factors such as levels of democracy, economic development, and electoral competitiveness in predisposing a state to the increased use of repression. Most notably, democratic forms of government—and democratization in general—has a pacifying effect on state repression [@DavenportArmstrong:2004; @Krain:1997; @Richards:1999]. Democracies have a negative effect on repression largely because of the increased probability of sanctions against authorities for undesirable behavior [@Davenport:2007a, 47].

More recent research focuses on how the behavior of various actors shapes government response. One of the most consistent findings in the political violence literature is that authorities repress to counter ongoing challenges to the status quo. In other words, dissent increases repressive behavior [@Davenport:2007]. However, dissent and repression have an endogenous relationship: state authorities also frequently engage in preventive repression to undermine or restrict groups' abilities to dissent [@RitterConrad:2016; @ConradRitter:2019]. Such preventive repression also damages organizations' capacity to impose costs [@Davenport:2007a, 47]. NGOs are one actor that can pose challenges to political authorities, and state repression of these actors, especially through legal restrictions can be considered a form of preventive repression [@Chaudhry:2016].

@@

Given this threat, a number of states decided to repress threatening NGOs. While violence is often effective at curbing dissent, it can also backfire, leading to widespread protests, decreasing leaders' legitimacy and increasing their criminal liability [@KimSikkink:2010; @Sikkink:2011]. In contrast to violence, administrative crackdown has fewer adverse consequences—citizens may see it as regulation rather than repression, resulting in little to no counter-mobilization. It may also circumvent concerns among state security or bureaucratic agents about illegitimate orders based on violent repression [@StephanChenoweth:2008]. There are also fewer international consequences since laws—unlike violence—rarely invite condemnation or threats of aid withdrawal [@CarothersBrechenmacher:2014].

This administrative crackdown via anti-NGO laws can provide convenient avenues for more violent repression and can foreshadow future government transgressions. Autocratic leaders want to avoid accountability and NGOs often stand in conflict with that goal. Civil society organizations, including NGOs, can help shape the narrative of leaders' performance both domestically and internationally. Monitoring by NGOs and subsequent shaming can include significant external revenue loss to states in the form of loss of aid, trade, and foreign direct investments [@KeckSikkink:1998; @BarryClayFlynn:2013]. Government attempts to repress NGOs can signal the beginning of a deteriorating human rights situation. Such repression reflects the government's willingness and capability to use measures that can make it avoid accountability. Once a state targets NGOs, over time it may grow less hesitant to apply violent repression more widely and more severely, thus leading to broader contractions of civil and human rights throughout the country.

@@

Simply noting the passage of (or counting) these laws may therefore miss the effect of their de facto implementation. Moreover, the implementation of these laws may not be limited to its legal character. The government may also engage in harassment of NGOs and activists to dissuade NGOs from carrying out their operations. This may include measures such as detentions, short-term incarceration, as well as beatings, threats to families and destruction of valuable property.

Finally, counting official laws can also undermeasure civil society repression because governments can repress NGOs without relying on formal laws. As prior research in Egypt, Ethiopia, and Russia points out, "rather than systematically enforcing all repressive legal measures-often impossible owing to capacity constraints," governments may choose to use extra-legal means to intimidate, harass or prosecute a few NGOs or activists. These cases, especially during moments or crises or contention, "highlight to the wider NGO community-and citizenry-that the government is willing and capable of using the law to repress dissent" [@Brechenmacher:2017, 94].

@@

(ref:chaudhry) @Chaudhry:2016

@@

To measure de jure repression, we use data on anti-NGO laws from @Chaudhry:2016 which includes official NGO laws for all countries (excluding micro-states) between 1990 and 2013. This measure of formal legal barriers does not necessarily represent individual laws. Instead, we count the presence or absence of specific types of anti-NGO legal barriers as a binary variable. If a country passes a broader NGO laws that implements several of these barriers simultaneously (e.g. making registration burdensome and restricting foreign funds), the values for each barrier will flip from 0 to 1. We create indices of the total number of legal barriers in each country based on three broad categories of laws: barriers to entry, funding, and advocacy. We also sum each of these indices to create an overall count of legal barriers.

@@

To measure de facto civil society repression in a country, we use the Varieties of Democracy's (V-Dem) CSO repression indicator (`v2csreprss` in @vdem-v10), which captures the extent to which governments harass, deter, liquidate, and arrest members of civil society organizations. This variable captures repression using an ordinal scale, defining civil society repression as severe, substantial, moderate, weak, or absent. This ordinal scale is then aggregated with a Bayesian item response theory measurement model and converted to a continuous scale that ranges from roughly −3 to 3, with higher values indicating less repression.

Figure \@ref(fig:barriers) shows each of these types of legal barriers per country over time, as well as average levels of de facto civil society repression by regime type. While anti-NGO laws have increased steadily since 1990—particularly laws related to foreign funding—there has not necessarily been a corresponding rise in de facto civil society repression, which has remained fairly constant across different regime types. This divergence is likely indicative of the split between de jure laws and their de facto implementation—a rise in laws is not necessarily tied to direct CSO repression.

@@

We present the results of our models in two stages. First, to see how minor shifts in civil society laws and repression influence future human rights, we examine the marginal effects of our main explanatory variables when all other model covariates are held constant. These effects are by no means causal—rather, they demonstrate how much variation in predicted human rights is possible as the legal environment for civil society improves or deteriorates within a typical country. Second, we present predictive diagnostics for each of our models to examine the improvements in predictions. The results of the ordered logistic regression models can be unwieldy to interpret with plain numbers due to varying intercepts and thresholds between PTS levels. Accordingly, we present graphical results for all our models wherever possible and include complete tables of model results in the appendix (see Tables A2 and A3). For ease in comparing the count of NGO laws with civil society repression, we reverse the x-axis in the marginal effects plots for our second empirical expectation so that increasing the civil society repression represents *worse* repression rather than less.

@@

In general, when holding all other covariates constant, the passage of an additional anti-NGO legal barrier is generally associated with a higher probability of more severe level of political terror in the following year. The probability of seeing different levels of political terror varies across different possible counts of NGO laws (see Figure \@ref(fig:plot-mfx-e1a)). In the absence of formal legal barriers, Level 2 of the PTS—an environment where political imprisonment, torture, and beatings are rare—is the most likely category of political terror across all models, appearing a predicted 60% of the time. Level 1, where a country is under secure rule of law, is the second most likely outcome at slightly under 40%. The other categories of the PTS are extremely rare when there are no anti-NGO laws—as such, we collapse them into a single category.

@@

As new laws are added, though, the distribution of probabilities for each of the PTS levels shifts. Adding new barriers to advocacy, entry, and funding decreases the likelihood of seeing Level 1 and increases the probability of a country turning to either Level 2 or Level 3 (with extensive political imprisonment and trial-free detention). The marginal effects of all forms of NGO laws results in this turn from Level 1 to Levels 2 or 3+, but this switching effect is strongest following additional barriers to advocacy. Anti-NGO laws do not predict more dramatic shifts in the PTS. Across all counts of legal barriers, the overall probability of seeing Levels 4 and 5 is minuscule, and the passage of additional laws does little to shift that probability. This is not unexpected however—the PTS has a natural ceiling at Level 5. Specifically, in addition to measuring the scope and intensity of government violence, the PTS scale also looks at the proportion of population targeted for abuse. Countries that receive a 4 and 5 are rare, as 4 denotes that murders, disappearances, and torture are a common part of life, while 5 denotes that physical abuses are routinely perpetrated against citizens not involved in politics, and includes cases where governments engage in "scorched earth" counterinsurgency policies or summary executions based on class or ethnic affiliation [@GibneyCornettWood:2019]. The majority of movement across PTS categories should thus occur between Levels 1–3.

Figure \@ref(fig:plot-mfx-e1b) shows the marginal effects of additional NGO legislation on latent human rights scores. Across all four models for different categories of legal barriers, holding all other values constant, the predicted level of human rights respect decreases. Adding a new NGO law is associated with a `r round(lhr_barriers_stats$barriers_total$estimate, 3)`-point decline in the level of respect for human rights in the following year, on average, representing a roughly 5% decline in predicted the predicted value. The marginal effect is larger (`r round(lhr_advocacy_stats$advocacy$estimate, 2)`, or a 9% decline) for additional barriers to advocacy and entry, which represent more burdensome restrictions than limits to funding.

@@

### How does general repression change following civil society repression?

@@

When holding all other factors equal, the de facto implementation of anti-civil society measures thus appear to be a stronger marginal indicator of future repression than the passage of formal de jure laws. In countries with low levels of political terror, it can potentially serve as a weak signal of a more general increase in political violence in the future, but predict more substantial changes in general respect for human rights. The difference in the effects of laws vs. their de facto implementation could also potentially be explained by the time frame of these strategies. As we explore in the next section, de jure civil society laws are a long-term strategy to prevent threats to the regime from coalescing in the first place. As such, the long-term effects of these laws in terms of their predictive power may require a much longer time period under study. In contrast, de facto crackdown on NGOs may signal the need to react sooner due to the immediacy of the threat.

@@

We assess predictive power in a few different ways, depending on the model used to predict the outcome. For the ordinal Political Terror Scale, Table \@ref(tab:correct-not-correct) shows a count of correct and incorrect predictions from each model. We measure the predicted accuracy of our latent human rights models with the root mean squared error, provided in Table \@ref(tab:tbl-rmse). In the appendix we include lists of country-year cases that see improved predictions when taking civil society laws and repression into account.

@@

# Discretionary disconnect between de jure and de facto NGO regulations: the case of Egypt

When looking both at marginal effects and improvements in predictive power, we thus find that the de facto implementation of anti-civil society measures are a stronger indicator of future repression than formal de jure laws. One reason for this disconnect between law and implementation is that laws targeted at NGOs are often not designed to be used immediately, but instead serve to protect regimes against potential threats in the future. Egypt provides a prime example of this mechanism.

@@

Law 84 (and its replacement Law 70 of 2017) is thus a useful lever for selectively controlling the allowable space for dissidence and leaving open the possibility of future human rights abuses. The law's built-in discretionary authority permits the government to give the appearance of concessions and openness to civil society while simultaneously providing the government with the ability to target individual NGOs and reserving the right to engage in broader crackdowns of associational rights. This trend holds across many other countries—anti-NGO laws can serve as a useful avenue for future human rights abuses, sometimes with years separating the official passage of the law and the large-scale implementation of the law.

>>>>

---
title: "NGO Repression as a Predictor of Worsening Human Rights Abuses"
short-title: "NGO Repression and Human Rights"
author:
- name: Suparna Chaudhry
  affiliation: Lewis & Clark College
  email: schaudhry@lclark.edu
  url: http://www.suparnachaudhry.com/
- name: Andrew Heiss
  affiliation: Georgia State University
  email: aheiss@gsu.edu
  url: https://www.andrewheiss.com/
date: "November 15, 2021"
published: "Conditionally accepted at the *Journal of Human Rights*"
code-repo: "Access the code and full analysis notebook at <https://stats.andrewheiss.com/canary-ngos/>"
abstract: >-
  An increasing number of countries have recently cracked down on non-governmental organizations (NGOs). Much of this crackdown is sanctioned by law and represents a bureaucratic form of repression that could indicate more severe human rights abuses in the future. This is especially the case for democracies, which unlike autocracies, may not aggressively attack civic space. We explore if crackdowns on NGOs predict broader human rights repression. Anti-NGO laws are among the most subtle means of repression and attract lesser domestic and international condemnation compared to the use of violence. Using original data on NGO repression, we test whether NGO crackdown is a predictor of political terror, and violations of physical integrity rights and civil liberties. We find that while de jure anti-NGO laws provide little information in predicting future repression, their patterns of implementation—or de facto civil society repression—predicts worsening respect for physical integrity rights and civil liberties.
# keywords: civil society, NGOs, repression, human rights, political terror, civil liberties
word-count: "8,007"
thanks: >-
  We would like to thank participants of the 2021 International Studies Association (ISA) convention, and the "Human Rights on the Edge: The Future of International Human Rights Law and Practice" workshop hosted by the Global Human Rights Hub at Arizona State University for their comments and feedback.
reference-section-title: References
bibliography: bibliography.bib
toc: false
title-page: true
endnotes: true
endfloat: false
mainfont: Cochineal
sansfont: Inter
author-note: >-
  Suparna Chaudhry, PhD, Assistant Professor of International Affairs, Department of International Affairs, Lewis \& Clark College; Andrew Heiss, PhD, Assistant Professor of Public Management, Andrew Young School of Policy Studies, Georgia State University. 
correspondence: >-
  Correspondence concerning this article should be addressed to Andrew Heiss, PhD, Department of Public Management and Policy, Andrew Young School of Policy Studies, Georgia State University, 14 Marietta Street NW, Suite 357, Atlanta, GA 30303. Contact: aheiss@gsu.edu
editor_options: 
  chunk_output_type: console
---

@@

Over the last few decades, governments across the globe have been increasingly cracking down on civil society organizations (CSOs), particularly non-governmental organizations (NGOs)—a phenomenon commonly known as closing or shrinking civic space [@CarothersBrechenmacher:2014; @DPR:2016; @Chaudhry:2016; @DupuyPrakash:2018]. More than 100 countries have proposed or enacted 244 different measures that have restricted, repressed, or shut down civil society since 2013 [@icnl:2021]. In Human Rights Watch's 2016 World Report, Executive Director Kenneth Roth argued that civil society was under more aggressive attack than at any time in recent memory. As seen in Figure \@ref(fig:civicus-map), by the end of 2020, only `r civicus$Open$prop_nice` of countries had open and unrestricted civic space [@civicusmonitor; @ChaudhryHeiss:2021].[^open]

@@

Much of this targeting of NGOs has been non-violent. Rather than publicly arrest or beat activists, states implement "administrative crackdown," or the passage of anti-NGO laws to create barriers to NGO advocacy, entry, and funding [@Chaudhry:2022].[^1] The use of legal tools instead of violent repression means that NGOs have to adapt their practices to stay within the bounds of permitted activities and operations [@Heiss:2019; @Heiss:2017]. While this can lead to the depoliticization and avoidance of direct political activism by CSOs [@BloodgoodTremblay-Boire:2017], states cannot soften or neutralize all forms of political activity and advocacy in subtle, legal ways like they have with NGOs. Sooner or later, states may need to deal with threats to their stability in a more direct, violent, confrontational manner. Anti-NGO laws could be an important early warning sign—or a canary in the coal mine—of worsening human rights repression, signaling that a state will continue down the path of further repression. What does state targeting of NGOs tell us about the state's broader respect for civil society and the overall level of human rights repression? Can the crackdown on NGOs act as a predictor of subsequent increasing repression?

In this paper, we explore the power of anti-NGO repression in predicting political terror and broader patterns of human rights abuses. We argue that information about NGO crackdown can be a valuable indicator of human rights trajectories. It can act as an early warning signal for worsening repression, suggesting which countries are testing the waters with comparatively "milder" methods of crackdown before enacting more violent forms of repression. To assess if attacks on NGOs and activists can provide us useful information about future repression, we use an original dataset on state crackdown on NGOs using legal repression across all countries from 1990–2013. We differentiate between two forms of NGO repression: de jure anti-NGO laws, and their de facto implementation. While the former captures anti-NGO laws that seek to repress these organizations, the latter captures their effects when they are actually implemented or enforced. The distinction is important—laws may frequently be passed to create a chilling effect on civil society. They provide a legal threat that could compel NGOs to self-censor, without actually seeing any uniform or widespread implementation.

Using multilevel Bayesian modeling, we find that formal de jure laws potentially provide a weak signal of future repression. Increasing de facto civil society repression, on the other hand, predicts both a moderate increase in the probability of moderate political terror and a deterioration in the overall respect for human rights in a typical country.

Our paper makes two important contributions to the fields of repression and international human rights law. First, in predicting the onset of repressive state policies, while structural factors such as levels of democracy, economic development, electoral competitiveness can identify countries that are likely to have more or less respect for civic freedoms, these factors are also slow to change over time. As such, they may not be particularly suitable for forecasting short-term trajectories, especially for moderately repressive countries. On the other hand, state repression of NGOs is a more dynamic process that may provide insight into changing levels of human rights standards and repression in subsequent years. While recent research has made important contributions to studying the extent and effects of state crackdown on NGOs, much of this focuses on de jure restrictions and its impact on NGOs and donors [@christensen2013; @DPR:2016; @DupuyPrakash:2018; @BromleySchoferLonghofer:2020; @BakkeMitchellSmidt:2020; @SmidtPereraMitchell:2021]. By looking at laws *and* their actual practices of implementation, we are able to assess the long-term effects of closing civic space on broader patterns of repression.

[^1]: In this article, we use the terms "anti-NGO laws" and "de jure NGO laws" interchangeably.

@@

Below, we first look at existing literature on repression and political violence, outlining state motivations to repress NGOs. We then elaborate on why attacking NGOs may be an important predictor of subsequent repression. Using both original data as well as data from several existing datasets, we build an ensemble of predictive models and explore (1) the marginal effects of civil society repression on different forms of broader human rights repression, and (2) the out-of-sample predictive performance of each model. We then illustrate one of the mechanisms of our findings—the distinct impact of de jure and de facto anti-NGO repression—with a brief case illustration of Egypt's treatment of civil society since 2002. We conclude by discussing the implications of our findings on autocratic survival, and how states may manage NGOs as a strategy to remain in power.

@@

Repression is the use of coercive action against an individual or organization, for the purpose of imposing a cost on the target as well as deterring specific activities perceived to be challenging to government personnel, practices, or institutions [@Goldstein:1978, p. xxvii; @Davenport:2007, 487; and @ConradRitter:2019, 15]. Early literature on repression focused on the role of structural factors such as levels of democracy, economic development, and electoral competitiveness in predisposing a state to the increased use of repression. Most notably, democratic forms of government—and democratization in general—has a pacifying effect on state repression [@DavenportArmstrong:2004; @Krain:1997; @Richards:1999]. Democracies have a negative effect on repression largely because of the increased probability of sanctions against authorities for undesirable behavior [@Davenport:2007a, 47].

More recent research focuses on how the behavior of various actors shapes government response. One of the most consistent findings in the political violence literature is that authorities repress to counter ongoing challenges to the status quo. In other words, dissent increases repressive behavior [@Davenport:2007]. However, dissent and repression have an endogenous relationship: state authorities also frequently engage in preventive repression to undermine or restrict groups' abilities to dissent [@RitterConrad:2016; @ConradRitter:2019]. NGOs are one actor that can pose challenges to political authorities, and state repression of these actors—especially through anti-NGO laws—can be considered a form of preventive repression [@Chaudhry:2022].

@@

Given this threat, a number of states decided to repress threatening NGOs. While violence is often effective at curbing dissent, it can also backfire, leading to widespread protests, decreasing leaders' legitimacy and increasing their criminal liability [@KimSikkink:2010; @Sikkink:2011]. In contrast to violence, anti-NGO laws have fewer adverse consequences—citizens may see it as regulation rather than repression. It may also circumvent concerns among state security or bureaucratic agents about illegitimate orders based on violent repression [@StephanChenoweth:2008]. There are also fewer international consequences since laws—unlike violence—rarely invite condemnation or threats of aid withdrawal [@CarothersBrechenmacher:2014].

These anti-NGO laws can provide convenient avenues for more violent repression and can foreshadow future government transgressions. Autocratic leaders want to avoid accountability and NGOs often stand in conflict with that goal. Civil society organizations, including NGOs, can help shape the narrative of leaders' performance both domestically and internationally. Monitoring by NGOs and subsequent shaming can include significant external revenue loss to states in the form of loss of aid, trade, and foreign direct investments [@KeckSikkink:1998; @BarryClayFlynn:2013]. Government attempts to repress NGOs through the use of law can signal the beginning of a deteriorating human rights situation. Such repression reflects the government's willingness and capability to use measures that can make it avoid accountability. Once a state targets NGOs, over time it may grow less hesitant to apply violent repression more widely and more severely, thus leading to broader contractions of civil and human rights throughout the country.

@@

Simply noting the passage of (or counting) these laws may therefore miss the effect of their de facto implementation. Some governments may even pass laws to create a chilling effect on these organizations. By providing a legal threat, they could compel these organizations to self-censor without even needing to implement these laws in any systematic or meaningful capacity. These laws, especially during moments or crises or contention, "highlight to the wider NGO community-and citizenry-that the government is willing and capable of using the law to repress dissent" [@Brechenmacher:2017, 94].

Moreover, the implementation of these laws may not be limited to their legal character. The government may also engage in harassment of NGOs and activists to dissuade NGOs from carrying out their operations. This may include measures such as detentions, short-term incarceration, as well as destruction of valuable property. Finally, counting official laws can also underestimate levels of NGO repression because governments can repress these organizations without relying on formal laws. As prior research in Egypt, Ethiopia, and Russia points out, "rather than systematically enforcing all repressive legal measures-often impossible owing to capacity constraints," governments may choose to use extra-legal means to intimidate, harass or prosecute a few NGOs or activists. These can include arrests, trials, imprisonment, and other violent sanctions against NGO activists such as beatings, threats to families, and attacks on property. As such, de facto crackdown in the form of not just legal harassment but also material and violent sanctions could lead to increases in both physical integrity and civil liberties violations.

@@

(ref:chaudhry) @Chaudhry:2022

@@

To measure de jure repression, we use data on anti-NGO laws from @Chaudhry:2022 which includes official NGO laws for all countries (excluding micro-states) between 1990 and 2013. This measure of formal legal barriers does not necessarily represent individual laws. Instead, we count the presence or absence of specific types of anti-NGO legal barriers as a binary variable. If a country passes a law that implements several of these barriers simultaneously (e.g. making registration burdensome and restricting foreign funds), the values for each barrier will flip from 0 to 1. We create indices of the total number of legal barriers in each country based on three broad categories of laws: barriers to entry, funding, and advocacy. We also sum each of these indices to create an overall count of legal barriers.

@@

To measure de facto civil society repression in a country, we use the Varieties of Democracy's (V-Dem) CSO repression indicator (`v2csreprss` in @vdem-v10), which captures the extent to which governments harass, deter, liquidate, and arrest members of civil society organizations. This variable captures repression using an ordinal scale, defining civil society repression as severe, substantial, moderate, weak, or absent. At the most severe levels, the government violently and actively pursues NGOs, seeking to not only deter their activities but also liquidate them. On the least repressive end of the spectrum, NGOs are free to organize, associate, express themselves and criticize the government without fear of government sanctions or harassment. This ordinal scale is then aggregated with a Bayesian item response theory measurement model and converted to a continuous scale that ranges from roughly −3 to 3, with higher values indicating less repression.

Figure \@ref(fig:barriers) shows each of these types of legal barriers per country over time, as well as average levels of de facto civil society repression by regime type. While anti-NGO laws have increased steadily since 1990—particularly laws related to foreign funding—there has not necessarily been a corresponding rise in de facto civil society repression, which has remained fairly constant across different regime types. This divergence is likely indicative of the split between de jure laws and their de facto implementation—new legislation is not necessarily implemented immediately and does not always lead to broader NGO repression.

@@

We present the results of our models in two stages. First, to see how minor shifts in anti-NGOs laws and de facto repression influence future human rights, we examine the marginal effects of our main explanatory variables when all other model covariates are held constant. These effects are by no means causal—rather, they demonstrate how much variation in predicted human rights is possible as the legal environment for civil society improves or deteriorates within a typical country. Second, we present predictive diagnostics for each of our models to examine the improvements in predictions. The results of the ordered logistic regression models can be unwieldy to interpret with plain numbers due to varying intercepts and thresholds between PTS levels. Accordingly, we present graphical results for all our models wherever possible and include complete tables of model results in the appendix (see Tables A2 and A3). For ease in comparing the count of NGO laws with civil society repression, we reverse the x-axis in the marginal effects plots for our second empirical expectation so that increasing the civil society repression represents *worse* repression rather than less.

@@

In general, when holding all other covariates constant, the passage of an additional anti-NGO law is generally associated with a higher probability of more severe level of political terror in the following year. The probability of seeing different levels of political terror varies across different possible counts of NGO laws (see Figure \@ref(fig:plot-mfx-e1a)). In the absence of formal laws, Level 2 of the PTS—an environment where political imprisonment, torture, and beatings are rare—is the most likely category of political terror across all models, appearing a predicted 60% of the time. Level 1, where a country is under secure rule of law, is the second most likely outcome at slightly under 40%. The other categories of the PTS are extremely rare when there are no anti-NGO laws—as such, we collapse them into a single category.

@@

As new laws are added, though, the distribution of probabilities for each of the PTS levels shifts. Adding new barriers to advocacy, entry, and funding decreases the likelihood of seeing Level 1 and increases the probability of a country turning to either Level 2 or Level 3 (with extensive political imprisonment and trial-free detention). The marginal effects of all forms of anti-NGO laws results in this turn from Level 1 to Levels 2 or 3+, but this switching effect is strongest following additional barriers to advocacy. Anti-NGO laws do not predict more dramatic shifts in the PTS. Across all counts of anti-NGO laws, the overall probability of seeing Levels 4 and 5 is minuscule, and the passage of additional laws does little to shift that probability. This is not unexpected however—the PTS has a natural ceiling at Level 5. Specifically, in addition to measuring the scope and intensity of government violence, the PTS scale also looks at the proportion of population targeted for abuse. Countries that receive a 4 and 5 are rare, as 4 denotes that murders, disappearances, and torture are a common part of life, while 5 denotes that physical abuses are routinely perpetrated against citizens not involved in politics, and includes cases where governments engage in "scorched earth" counterinsurgency policies or summary executions based on class or ethnic affiliation [@GibneyCornettWood:2019]. The majority of movement across PTS categories should thus occur between Levels 1–3.

Figure \@ref(fig:plot-mfx-e1b) shows the marginal effects of additional anti-NGO laws on latent human rights scores. Across all four models for different categories of legal barriers, holding all other values constant, the predicted level of human rights respect decreases. Adding a new anti-NGO law is associated with a `r round(lhr_barriers_stats$barriers_total$estimate, 3)`-point decline in the level of respect for human rights in the following year, on average, representing a roughly 5% decline in predicted the predicted value. The marginal effect is larger (`r round(lhr_advocacy_stats$advocacy$estimate, 2)`, or a 9% decline) for additional barriers to advocacy and entry, which represent more burdensome registration requirements than barriers to funding.

@@

### How does general repression change following de facto civil society repression?

@@

When holding all other factors equal, the de facto implementation of anti-NGO laws thus appear to be a stronger marginal indicator of future repression than simply the passage of these laws. While these laws may be intended to have a chilling effect on civil society (and certainly have a chilling effect on levels of aid and NGO operations as shown by recent literature), these de jure laws by themselves do not have similar adverse consequences on respect for physical integrity rights and civil liberties. In countries with low levels of political terror, it can potentially serve as a weak signal of a more general increase in political violence in the future, but predict more substantial changes in general respect for human rights. The difference in the effects of laws vs. their de facto implementation could also potentially be explained by the time frame of these strategies. As we explore in the next section, de jure civil society laws are a long-term strategy to prevent threats to the regime from coalescing in the first place. As such, the long-term effects of these laws in terms of their predictive power may require a much longer time period under study. In contrast, de facto crackdown on NGOs may signal the need to react sooner due to the immediacy of the threat.

@@

We assess predictive power in a few different ways, depending on the model used to predict the outcome. For the ordinal PTS, Table \@ref(tab:correct-not-correct) shows a count of correct and incorrect predictions from each model. We measure the predicted accuracy of our latent human rights models with the root mean squared error, provided in Table \@ref(tab:tbl-rmse). In the appendix we include lists of country-year cases that see improved predictions when taking civil society laws and repression into account.

@@

# Discretionary disconnect between de jure and de facto NGO repression: The case of Egypt

When looking both at marginal effects and improvements in predictive power, we find that the de facto implementation of anti-civil society measures are a stronger indicator of future repression than de jure anti-NGO laws. One reason for this disconnect between law and implementation is that laws targeted at NGOs are often not designed to be used immediately, but instead serve to protect regimes against potential threats in the future. Egypt provides a prime example of this mechanism.

@@

Law 84 (and its replacement Law 70 of 2017) is thus a useful lever for selectively controlling the allowable space for dissidence and leaving open the possibility of future human rights abuses. The law's built-in discretionary authority permits the government to give the appearance of concessions and openness to civil society while simultaneously provides the government the ability to target individual NGOs and reserving the right to engage in broader crackdowns of associational rights.

This trend holds across many other countries as well. For instance, even though India passed an anti-NGO law (containing multiple bariers to funding and advocacy) in 1976, it has not always been uniformly implemented. Rather, having an anti-NGO law on the books serves the state's long-term strategic interests, enabling it to crack down on NGOs when they start posing electoral or economic threats to the state [Chaudhry:2022]. Anti-NGO laws can therefore serve as a useful avenue for future human rights abuses, sometimes with years separating the official passage of the law and the large-scale implementation of the law, explaining why de facto levels of NGO repression serve as a stronger predictor of worsening respect for human rights abuses.
